---
title: "project_2"
author: "Tanmay_Naik"
date: "2024-11-20"
output: html_document
---

```{r setup, include=FALSE}
#Yesknitr::opts_chunk$set(echo = TRUE)
```

HEATMAP:

Using BrailleR:

```{r}
library(tidyverse)
library(BrailleR)
library(ggplot2)
library(reshape2)
library(dplyr)
```

```{r}
data <- read.csv("https://raw.githubusercontent.com/INFO-526-FA24/project-2-4x4/refs/heads/main/data/world_happiness_report_2019.csv")
```

```{r}
colnames(data)
```

```{r}
# Select numerical variables for correlation analysis
selected_data <- data %>%
  select(
    "GDP.per.capita", 
    "Social.support", 
    "Healthy.life.expectancy", 
    "Freedom.to.make.life.choices", 
    "Generosity", 
    "Perceptions.of.corruption"
  )

# Calculate the correlation matrix
cor_matrix <- cor(selected_data, use = "complete.obs")

# Transform correlation matrix for heatmap plotting
tran_cor_matrix <- melt(cor_matrix)
```

```{r}
# Heatmap using BrailleR
heatmap_plot <- ggplot(data = tran_cor_matrix, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
                       midpoint = 0, limit = c(-1, 1), space = "Lab", 
                       name = "Correlation") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  labs(title = "Correlation Heatmap", x = "Attributes", y = "Attributes")

# Print the heatmap with BrailleR
VI(heatmap_plot)
```

```{r}
# Calculate and print a summary of the strongest and weakest correlations
summary_text <- function(cor_matrix) {
  strongest <- which(cor_matrix == max(cor_matrix[cor_matrix < 1]), arr.ind = TRUE)
  weakest <- which(cor_matrix == min(cor_matrix), arr.ind = TRUE)
  cat("Strongest correlation:", rownames(cor_matrix)[strongest[1]], "and",
      colnames(cor_matrix)[strongest[2]], "=", max(cor_matrix[cor_matrix < 1]), "\n")
  cat("Weakest correlation:", rownames(cor_matrix)[weakest[1]], "and",
      colnames(cor_matrix)[weakest[2]], "=", min(cor_matrix), "\n")
}

# Print the summary
summary_text(cor_matrix)
```

Using regular ggplot2:

```{r}
# Create the heatmap
heatmap_plot <- ggplot(data = tran_cor_matrix, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
                       midpoint = 0, limit = c(-1, 1), space = "Lab", 
                       name = "Correlation") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  labs(title = "Correlation Heatmap", x = "Attributes", y = "Attributes")

print(heatmap_plot)

```

The BrailleR package provides basic information about the plot such as the title, axis and legend description. It explains the range of correlation values and their mapping to colors, giving some sense of the data. But the VI functionality of BrailleR is not capable of processing the tile based data in heatmaps. It only summarizes the metadata and omits the main information. This limitation arises because heatmap is composed of tiles with colors representing values. The VI() function cannot generate meaningful descriptions for this style of visualization because the visual-to-text mapping isn't straightforward.Manual intervention is required to write the supplementary code to extract insights such as strongest and weakest correlations. This makes BrailleR inconvenient for heatmaps, as it does not fully represent the information on the plot in textual format for the visually impaired.

```{r}
#Bar Plot

top_bottom_countries <- data %>%
  arrange(desc(Score)) %>%           
  slice_head(n = 10) %>%              
  bind_rows(data %>%
              arrange(Score) %>%       
              slice_head(n = 10))

bar_plot <- ggplot(top_bottom_countries, aes(x = reorder(Country.or.region, Score), y = Score)) +
  geom_bar(stat = "identity", fill = "darkolivegreen4") +
  coord_flip() + 
  labs(title = "Happiness Index Score",
       subtitle = "Top 10 and Bottom 10 Ranking Countries",
       x = "Country",
       y = "Happiness Score") +
  scale_y_continuous(
    breaks = seq(0, 8, by = .5),
    expand = c(0, 0)
  ) +
  theme_minimal() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank()
  )

print(bar_plot)

```

Once again, the BrailleR package provides basic information about the plot such as the title, subtitle, axis and coloring. It says what countries are being plotted, but not what the specific values of the plot have, meaning that while a user might know what the plot is displaying and how, they wouldn't know what the actual data is. In addition, the descriptions of the format of the plot are confusing. BrailleR successfully recognizes that the axis are flipped, meaning that the x-axis is on the veritcal axis and the y is on the horizontal, but its language and descriptions switch between the standard display and the flipped display.

```{r}
#Pie Chart

library(RColorBrewer)

data <- data.frame(
  category = c("Under 5", "5-17", "18-24", "25-34", "35-44", "45-64", "65-84", "85 and over"),
  value = c(5.6, 16.5, 9.4, 13.5, 12.7, 25.4, 14.9, 1.91)
)

colors <- brewer.pal(length(unique(data$category)), "Set2")

data$category <- factor(data$category, levels = data$category)

data$fraction = data$value / sum(data$value)
data$percentage = round(data$fraction * 100, 1)

ggplot(data, aes(x = "", y = value, fill = category)) +
  geom_bar(stat = "identity", width = 1, color = "black") +
  coord_polar(theta = "y") +
  labs(title = "Age Groups in the United States", fill = "Category",
       caption="According to the 2020 US Census") +
  geom_text(aes(label = ifelse(percentage >= 6, paste(percentage, "%"), "")),
            position = position_stack(vjust = .5),
            color = "white", size = 5) +
  scale_fill_manual(values = colors) +
  theme_void() 

```

BrailleR does not properly portray pie charts for users. It provides information on the title, subtitle, and legend as well as describes the colors of each of the segments. However, the text description does not properly explain what the data is or what the data is showing. Reading the text description, it deeply goes into the construction of the plot and more technical aspects of how it was constructed which does not give blind users access to what the plot is showing.

```{r}
Gold_data <- read.csv("https://raw.githubusercontent.com/INFO-526-FA24/project-2-4x4/refs/heads/main/data/Gold%20%20Prices.csv")
```

```{r}
colnames(Gold_data)
```

```{r}
#plotting the closing price of gold over time with ggplot
Gold_data$Date <- as.Date(Gold_data$Date)

# Plotting the closing price over time
Timeseries_plot <- ggplot(Gold_data, aes(x = Date, y = Close)) +
  geom_line(color = "goldenrod", size = 1) +
  labs(title = "Gold Prices Over Time", x = "Date", y = "Closing Price (USD)") +
  theme_minimal()
print(Timeseries_plot)
```

```{r}
# Braille interpritation of the time series plot
VI(Timeseries_plot)
```

```{r}
# Calculate trends in the data (e.g., overall increase/decrease)
gold_summary <- Gold_data %>% 
  summarize(
    Start_Price = first(Close),
    End_Price = last(Close),
    Average_Price = mean(Close),
    Min_Price = min(Close),
    Max_Price = max(Close),
    Price_Change = last(Close) - first(Close)
  )

# Generate a verbal interpretation
VI(gold_summary)
```

For Time Series plot the BrailleR fails to interpret even the basic aspects of the plot such as the x-axis, though the interpretation of title, layer and color with the number of total observations. By creating the custom summary function and summarizing it using BrailleR provides the statistical details about the variables such start_price, End_price, Average_price and other, this only goes deep into statistical details which are not useful enough to give blind users access to what the plot is showing.

```{r}
# Regression plot of GDP per capita vs happiness score
regression_plot <- ggplot(data, aes(x = GDP.per.capita, y = Score)) +
  geom_point() +
  geom_smooth(method = "lm", color = "green") +
  labs(title = "Happiness Score vs. GDP per Capita", x = "GDP per Capita", y = "Happiness Score")

print(regression_plot)
```

```{r}
VI(regression_plot)
```

```{r}
# Fit a linear regression model
regression_model <- lm(Score ~ GDP.per.capita, data = data)

# Generate a textual summary using BrailleR
VI(summary(regression_model))
```

When used BrailleR for regression_plot, It provides basic information about the plot such as the title, axis, layer information and coloring. While used with underlying linear model that represents the plot (regression_model) provides statistical details like significance level, estimate and p-value. Here the significance of 1% represents strong statistical significanc, suggesting that the relationship is highly unlikely to be due to random chance. Estimate value represents the slope of the regression line. For every 1 unit increase in GDP per capita, the Happiness Score is expected to increase by 2.22 units on average.A positive estimate means there is a positive relationship: higher GDP per capita is associated with higher happiness scores. The p-value is extremely small, close to 0, which indicates a very strong likelihood that this relationship is real and not due to random variation in the data.

```{r}
# Calculate daily price changes
Gold_data <- Gold_data %>%
  mutate(Daily_Change = Close - lag(Close, default = first(Close)))

# Plot the histogram of daily price changes
gold_histogram <- ggplot(Gold_data, aes(x = Daily_Change)) +
  geom_histogram(binwidth = 5, fill = "gold", color = "black") +
  labs(title = "Histogram of Daily Gold Price Changes",
       x = "Daily Price Change (USD)",
       y = "Frequency") +
  theme_minimal()

print(gold_histogram)
```

```{r}
VI(summary(Gold_data$Daily_Change))
```

```{r}
VI(gold_histogram)
```
For the Histogram, using BrailleR summary function provides the statistical details about the Daily change in gold_prices which highlights The largest negative daily change as 10.209$, 25% of daily changes are below -0.82 USD, The middle value, indicating half of the daily changes are above and half are below 0.10 USD, On average, daily price changes are slightly positive and The largest positive daily change is 7.09 USD. While the VI(gold_histogram) function provides basic information about such as title, axis, color and details about the bars. 